Predicting Customer Behavior Using a Portuguese Financial Institution Dataset 

### 2024-07-13 code/eda-code/eda_my version.qmd
MAJOR CHANGES

# Setup
Necessary libraries for this script.
```{r}
#| echo: false
#| output: false
library(here) 
library(dplyr)
library(skimr)
library(ggplot2)
library(scales)
library(tidyr)
library(forcats)
library(corrplot)
```

@@

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "summarytable1.rds")
saveRDS(summary_df, file = summarytable_file)
```

@@

Trying to explore some additional aspects of the data

This shows the relationship between age and balance, grouped by education
```{r}
ggplot(data=mydata)+
  geom_point(mapping =aes(x=age, y=balance))+
  facet_wrap(~education, nrow=2)

Next, we can assess how age relates the number of jobs, grouped by job type.
```{r}
ggplot(data=mydata)+
  geom_bar(mapping =aes(x=age))+
  facet_wrap(~job, nrow=2)
```

Digging into the Duration variable, we can see some very extreme values affecting this variable
```{r}
ggplot(data=mydata)+
  geom_bar(mapping =aes(x=duration))

boxplot(mydata$duration)

@@

Taking a look at frequency counts across different variables of interest
```{r}
hist(numericvar$age)
hist(numericvar$balance)
hist(numericvar$duration)
hist(numericvar$campaign)
hist(numericvar$pdays)
hist(numericvar$previous)
```

@@

We've analyzed which job types had the MOST positive outcomes, but now we want to 
compare which job type had the greates PERCENT of positive outcomes.

@@

p13 = ggplot(pctjob, aes(x=reorder(job, numofYES), y=numofYES, fill=numofYES))+
  geom_bar(stat="identity")+
  coord_flip()

@@

>>>>

# Setup
Necessary libraries for this script.
```{r}
#| echo: false
#| output: false
library(here) 
library(dplyr)
library(skimr)
library(ggplot2)
library(scales)
library(tidyr)
library(forcats)
library(corrplot)
library(corrplot)
```

@@

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)

# save to file
summarytable_file = here("results", "tables", "summarytable1.rds")
saveRDS(summary_df, file = summarytable_file)
```

```{r}
library(knitr)
library(kableExtra)

data_location <- here::here("data","processed-data","processeddata1.rds")
mydata1 <- readRDS(data_location)

mydata1 = mydata1 %>% 
  select(age, job, marital, education, default, balance, housing, loan, contact, day, month, y_termSubscribed)

str_summary <- capture.output(str(mydata1))

describetbl1 = kable(str_summary, caption = "Summary of predictors") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F)

summarytable_file = here("results", "tables", "describetbl1.rds")
saveRDS(describetbl1, file = summarytable_file)

describetbl1
```

@@

@@

Taking a look at frequency counts across different variables of interest
```{r}
hist(mydata$age)
hist(mydata$balance)
```

@@

We've analyzed which job types had the MOST positive outcomes, but now we want to 
compare which job type had the greatest PERCENT of positive outcomes.

@@

```{r}
p13 = ggplot(pctjob, aes(x=reorder(job, numofYES), y=numofYES, fill=numofYES))+
  geom_bar(stat="identity")+
  coord_flip()

@@

```{r}
data_location <- here::here("data","processed-data","processeddata5.rds")
mydata <- readRDS(data_location)

numericonly = mydata %>% 
  select(!y_termSubscribed)

# Specify the file path where the image will be saved
file_path <- here("results", "figures","corplot.png")

# Open a PNG graphics device
png(filename = file_path, width = 800, height = 600)

# Create the correlation plot
corrplot(cor(numericonly))

# Close the graphics device
dev.off()
```

### 2024-07-13 code/processing-code/processingfile-v1 - my version.qmd
MAJOR CHANGES

We are mainly converting character variables into factors so we can do some categorical analysis.
I had initially included converting months to a numeric variable but realized that it was fine as a factor variable.
```{r}
d1 = rawdata |> 
  mutate(job = as.factor(job)) |>
  mutate(marital = as.factor(marital)) |>
  mutate(education = as.factor(education)) |>
  mutate(default = as.factor(default)) |>
  mutate(housing = as.factor(housing)) |>
  mutate(loan = as.factor(loan)) |>
  mutate(education = as.factor(education)) |>
  mutate(contact = as.factor(contact)) |>
  mutate(month = as.factor(month)) |>
  mutate(poutcome = as.factor(poutcome)) |>
  mutate(day = as.factor(day)) |>
  mutate(y = ifelse(y == "yes", 1, 0))

colnames(d1)[17] <- "y_termSubscribed"

@@

>>>>

We are mainly converting character variables into factors so we can do some categorical analysis. I had initially included converting months to a numeric variable but realized that it was fine as a factor variable.

```{r}
d1 = rawdata |> 
  mutate(job = as.factor(job)) |>
  mutate(marital = as.factor(marital)) |>
  mutate(education = as.factor(education)) |>
  mutate(default = as.factor(default)) |>
  mutate(housing = as.factor(housing)) |>
  mutate(loan = as.factor(loan)) |>
  mutate(education = as.factor(education)) |>
  mutate(contact = as.factor(contact)) |>
  mutate(month = as.factor(month)) |>
  mutate(y = ifelse(y == "yes", 1, 0)) |>
  mutate(y = as.factor(y)) |>
  mutate(poutcome = as.factor(poutcome)) |>
  mutate(day = as.factor(day)) |>
  mutate(y = as.factor(y)) %>% 
  select(!duration)

colnames(d1)[16] <- "y_termSubscribed"

@@

# Attempting to convert categorical variables to dummy

```{r}
library(caret)

testing_dummy = processeddata1 %>% 
  select(job, marital, education, default, housing, loan, contact) %>% 
  mutate(ID = row_number())

dummy <- dummyVars("~.", data=testing_dummy)
df1 <- predict(dummy, newdata = testing_dummy)

df1 = as.data.frame(df1)

numeric_vars = processeddata1 %>% 
  select(age, balance, day, y_termSubscribed) %>% 
  mutate(ID = row_number())

d2 <- merge(df1,numeric_vars, by = "ID")

```

```{r}
save_data_location <- here::here("data","processed-data","processeddata2.rds")
saveRDS(d2, file = save_data_location)
```

# Attempting to convert categorical variables to numeric factors

```{r}
library(caret)

d3 = processeddata1 |>
  select(job, marital, education, default, housing, loan, age, balance, day, y_termSubscribed) |>
  mutate(job = as.factor(unclass(job))) |>
  mutate(marital = as.factor(unclass(marital))) |>
  mutate(education = as.factor(unclass(education))) |>
  mutate(default = as.factor(unclass(default))) |>
  mutate(housing = as.factor(unclass(housing))) |>
  mutate(loan = as.factor(unclass(loan)))

summary(d3$education)
```

```{r}
save_data_location <- here::here("data","processed-data","processeddata3.rds")
saveRDS(d3, file = save_data_location)
```

# Attempting to convert categorical variables to numeric

```{r}
library(caret)

d4 = processeddata1 |>
  select(job, marital, education, default, housing, loan, age, balance, day, y_termSubscribed) |>
  mutate(job = unclass(job)) |>
  mutate(marital = unclass(marital)) |>
  mutate(education = unclass(education)) |>
  mutate(default = unclass(default)) |>
  mutate(housing = unclass(housing)) |>
  mutate(loan = unclass(loan))|>
  mutate(day = unclass(day))

```

```{r}
save_data_location <- here::here("data","processed-data","processeddata4.rds")
saveRDS(d4, file = save_data_location)
```

# Attempting to convert categorical variables to true numeric

```{r}
library(caret)

d5 = processeddata1 |>
  select(job, marital, education, default, housing, loan, age, balance, day, y_termSubscribed) |>
  mutate(job = as.numeric(job)) |>
  mutate(marital = as.numeric(marital)) |>
  mutate(education = as.numeric(education)) |>
  mutate(default = as.numeric(default)) |>
  mutate(housing = as.numeric(housing)) |>
  mutate(loan = as.numeric(loan))|>
  mutate(day = as.numeric(day))

```

```{r}
save_data_location <- here::here("data","processed-data","processeddata5.rds")
saveRDS(d5, file = save_data_location)
```

### 2024-07-13 products/manuscript/Manuscript.qmd
MAJOR CHANGES

---
title: "Working Title: Predicting Customer Behavior Using a Portuguese Financial Institution Dataset "
subtitle: ""
author: Antonio Flores
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/apa.csl
---

```{r, echo=FALSE, message=FALSE}
library(here)
library(knitr)
```

# Summary/Abstract
This project will seek to identify a model for predicting consumer financial behavior using a dataset from a Portuguese Bank.

@@

## General Background Information
Providing businesses with a model that will allow prioritization of consumers could save time and allow for a more accurate use of resources.

## Description of data and data source
The data was donated on 2/13/2012. It was collected from phone call marketing campaigns performed by a Portuguese banking institution.I have accessed this data from the UC Irvine Machine Learning Repository.

There are 45,212 records, and includes: age, marital status, job, education,details related to the phone call, as well as answers related to questions about past credit history. Additionally, the classification variable is whether or not the person subscribed to a term deposit. There are 17 features in total.

@@

I want to first perform some EDA to identify if there are any variables with extreme variation or conversely if there some patterns that can be seen already. Then I will test for correlation between different features and proceed from there.

## Schematic of workflow

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
```

## Data aquisition
Data retrieved from UCI ML Repository

@@

```{r}
#| label: tbl-summarytable1
#| tbl-cap: "Data Snapshot"
#| echo: FALSE
data_location <- here::here("data","raw-data","bank-full.csv")
raw=read.csv(data_location, header = TRUE, sep=";")
head(raw)
```

### Dimensions:
dim(raw)

```{r}
#| label: tbl-summarytable2
#| tbl-cap: "Data Dimensions"
#| echo: FALSE
dim(raw)
```

### Describing raw data
str(raw)

```{r}
#| label: tbl-summarytable3
#| tbl-cap: "Data Description"
#| echo: FALSE
str(raw)
```

## Statistical analysis

@@

@fig-result shows a boxplot figure comparing balance levels across job types.

```{r}
#| label: fig-result
#| fig-cap: "Job type and bank account balance"
#| echo: FALSE
knitr::include_graphics(here("results","figures","job-balance-scatter.png"))
```

@fig-result1 shows a bar chart showing education levels of the data.

```{r}
#| label: fig-result1
#| fig-cap: "Education level"
#| echo: FALSE
knitr::include_graphics(here("results","figures","education-barchart.png"))
```

@fig-result2 shows a scatter plot figure produced by one of the R scripts.

```{r}
#| label: fig-result2
#| fig-cap: "Age and bank account balance stratified by marital status"
#| echo: FALSE
knitr::include_graphics(here("results","figures","age-balance-stratified.png"))
```

@@

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
#resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
#knitr::kable(resulttable2)
```

## Full analysis

@@

## Summary and Interpretation

## Strengths and Limitations

## Conclusions

>>>>

---
title: "Predicting Customer Behavior Using a Portuguese Financial Institution Dataset "
subtitle: ""
author: Antonio Flores
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/apa.csl
---

```{r, echo=FALSE, message=FALSE, warning = FALSE}
library(here)
library(knitr)
```

# Summary/Abstract

This project will seek to identify a model for predicting consumer financial behavior using a dataset from a Portuguese Bank. After the data has been cleaned and prepared for analysis, exploratory data analysis will be performed to gather more information regarding the shape, size, and behavior of different variables, to gauge their usefulness in a prediction model. Additionally, several different statistical tests will be implemented to identify which classes within different variables affect the outcome variable. Finally this project will test different machine learning models and conclude with the resulting findings.

@@

## General Background Information

Providing businesses with a model that will allow prioritization of consumers and/or demographics has great potential in improving resource management and future marketing campaigns, as well as increasing efficient spending. This particular project will utilize Classification Prediction. Unlike Regression Prediction which will attempt to predict a continuous value (e.g., x amount of dollars, x amount of cells), Classification Prediction seeks to train a model that can correctly predict a classification (e.g., True/False, Success/Failure) based on the given predictors. In our case, our y variable or response variable is a binary variable, Yes/No, answering whether or not a customer subscribed to a term deposit.

## Description of data and data source

The data was donated on 2/13/2012. It was collected from phone call marketing campaigns performed by a Portuguese banking institution.I have accessed this data from the UC Irvine Machine Learning Repository.

There are 45,212 records,17 columns/variables which include: age, marital status, job, education,details related to the phone call, as well as answers related to questions about past credit history. Additionally, as mentioned, the classification variable is whether or not the person subscribed to a term deposit.

@@

## Data aquisition

The dataset for this project was retrieved from UCI ML Repository in CSV form. Additionally, I created a codebook based on data from the same source.

@@

```{r}
#| label: tbl-summarytable1
#| tbl-cap: "Data Snapshot"
#| echo: FALSE
data_location <- here::here("data","raw-data","bank-full.csv")
raw=read.csv(data_location, header = TRUE, sep=";")
kable(head(raw[1:5]))

kable(head(raw[6:10]))
```

### Dimensions:

Rows: 45, 211\
Columns: 17

### Describing data

```{r, echo=FALSE, message=FALSE, warning = FALSE}
data_location <- here::here("results","tables","describetbl1.rds")
describe1 <- readRDS(data_location)
describe1
```

### Cleaning Data

There were several different data cleaning methods that were experimented with in order to find the best way to prepare the data for modeling. First, I converted several variables to factors as they had been read in as character variables, or strings. Next I tried to use the DummyVars tool to convert every categorical variable to a dummy value but this made modeling difficult due to a large amount of binary variables. Instead I proceeded with converting the categorical variables to numeric values while maintaining their factor status. I also created a dataset with these numeric values that were not factors. These were the two primary dataset I used for my modeling. Finally I created a dataset stripping the categorical values of their attributes, this was solely created to allow for my corrplot to work.

### Removing Initial Predictors

I removed the 'Duration' variable due to a warning I discovered on the site hosting this dataset initially: "Important note: this attribute highly affects the output target (e.g., if duration=0 then y='no'). Yet, the duration is not known before a call is performed. Also, after the end of the call y is obviously known. Thus, this input should only be included for benchmark purposes and should be discarded if the intention is to have a realistic predictive model."(@misc_bank_marketing_222) Due to this new information I decided to remove all the variables related to this, which included 'campaign', 'pdays', 'previous', and 'poutcome.' Bringing our new total number of predictors to 11.

@@

In this section we will be exploring the data using different charts to identify outliers, abnormalities, relationships, and the general shape and feel of different variables.

@fig-result1 shows the distribution of the 'Age' variable.

```{r}
#| label: fig-result1
#| fig-cap: "Age Histogram"
#| echo: FALSE
knitr::include_graphics(here("results","figures","age-distribution.png"))
```
While this shows a slight skew to the left, if we account for the extreme value on the right, this data seems to be fairly normally distributed.

@fig-result2 shows the distribution of the 'Education' variable

```{r}
#| label: fig-result2
#| fig-cap: "Education level Bar Chart"
#| echo: FALSE
knitr::include_graphics(here("results","figures","education-barchart.png"))
```

@fig-result3 shows the most common 'Job' types in our data

```{r}
#| label: fig-result3
#| fig-cap: "Most Common Job Types"
#| echo: FALSE
knitr::include_graphics(here("results","figures","job-barchart.png"))
```

Interestingly, Blue-collar and Management are the two most common job types. This play an important role later in our analysis due to the percent of the data these two categories represent.

@fig-result4 shows a scatter plot figure produced by one of the R scripts.

```{r}
#| label: fig-result4
#| fig-cap: "Age and bank account balance stratified by marital status"
#| echo: FALSE
knitr::include_graphics(here("results","figures","age-balance-stratified.png"))
```

@fig-result5 shows a barplot of most common days of the month to record a positive outcome.

```{r}
#| label: fig-result5
#| fig-cap: "Days of the month for positive outcomes"
#| echo: FALSE
knitr::include_graphics(here("results","figures","days-yes.png"))
```
While there doesn't appear to be a significant trend, we can clearly see that the 30th of the month stands out as being a day of interest, especially compared to other days of the months.

@fig-result6 shows potential correlation between different variales.

```{r}
#| label: fig-result6
#| fig-cap: "Correlation Matrix"
#| echo: FALSE
knitr::include_graphics(here("results","figures","corplot.png"))
```

The biggest takeaways from our Exploratory Data Analysis is that most variables are fairly normally distributed due to the size of the data, and most variable are not correlated to each other, with one exception ('Age' and 'Marital')

@@

Before we begin with the machine learning analysis, I sought to test a few variables of interest ('Age', 'Job') in their significance of affecting the response variable. As mentioned previously, our response variable is a binary value (Yes/No), which means that we need to use a logistic regression model instead of a linear regression model.

Below are the results for our first logistic model fit with Age as Predictor.

```{r}
#| label: basicmodel1
#| fig-cap: "Logistic Regression with Age as Predictor"
#| echo: FALSE

data_location <- here::here("results","tables","basicmodel1.rds")
basic1 <- readRDS(data_location)
kable(basic1)
```

Below are the results for our second logistic model fit with Job as Predictor.

```{r}
#| label: basicmodel2
#| fig-cap: "Logistic Regression with Job as Predictor"
#| echo: FALSE

data_location <- here::here("results","tables","basicmodel2.rds")
basic2 <- readRDS(data_location)
kable(basic2)
```

As expected, both predictors are very significant. We will proceed to the Machine Learning portion of this project.

## Full analysis
So far, I have focused on the MARS model, KNN, and Logistic Regression to determine the prediction model for this project. As highlighted in the data cleaning section, there were two datasets I was focusing on. One that had numeric factors for the categorical predictors, and one that had just numeric variables for the categorical predictors. I decided to examine both so there are a total of 6 different runs recorded below, 3 runs for each dataset.

@tbl-resulttable1 displays the Accuracy and Kappa scores for the first three model runs using the predictors that are numeric and are not factors.
```{r}
#| label: tbl-resulttable1
#| tbl-cap: "Performance Results for Models using Numeric (unfactored) Predictors"
#| echo: FALSE
resulttable1 = readRDS(here("results","tables","perftable1.rds"))
knitr::kable(resulttable1)
```

@tbl-resulttable2 displays the Accuracy and Kappa scores for the first three model runs using the predictors that are numeric and are not factors.
```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Performance Results for Models using Numeric Factored Predictors"
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","perftable2.rds"))
knitr::kable(resulttable2)
```

In terms of accuracy, all 6 model runs seem to have almost identical accuracy. They do tend to differ quite a bit in terms of Kappa...

Finally, we can examine which predictors specifically improved prediction the most.

The following figures were taken from the Factored Numeric model runs.

@fig-result7 shows the Predictors that had the most impact on the Mars Model

```{r}
#| label: fig-result7
#| fig-cap: "Mars Model Variables of Importance"
#| echo: FALSE
knitr::include_graphics(here("results","figures","importantMM2.png"))
```

Housing2 indicates that the person does have a housing loan.\
Loan2 indicates that the person does have a personal loan for an unspecified purpose.\
Job9 indicates that the person is a Student.\
education 3 indicates that the person has attained Tertiary Education.\

@fig-result8 shows the Predictors that had the most impact on the KNN Model

```{r}
#| label: fig-result8
#| fig-cap: "KNN Variables of Importance"
#| echo: FALSE
knitr::include_graphics(here("results","figures","importantKNN2.png"))
```

@@

## Summary and Interpretation
So far all of the models I have chosen have shown considerable predictive ability. I plan to tune the existing three models while also trying out Random Forest as well as potentially one more model. 
## Strengths and Limitations

## Conclusions
I will conclude with choosing the model that seems the most robust, accurate, and flexible, as well as determining which variables have the greatest predictive ability and which levels within those predictors stand out as particularly effective in classifying correctly.
